---
title: "RGCA_Manuscript_Pipe" author: "Daniel Zilber" date: "2023-07-07" output:
pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#, echo=FALSE
```
## The Mixture Prediction Pipeline
This script predicts a response to a mixture of chemicals given an input data of
individual dose responses and the concentrations of the components of the
mixture.  The procedure followed in this pipeline is as follows:
1. Load all inputs 2. Fit individual dose response curves 3. Cluster the
individual dose response slope parameters 4. Create a collection of predictors
("calculators") by sampling the slope cluster and parameters 5. Given a mixture,
predict a response with each calculator.  The plots show the 5, 50, and 95
percentile.
### Loading data To start, we load the required packages, files with helper
functions or MCMC scripts, and a script that parses the input data.
```{r Input}
library(ggplot2)
library(cowplot)
library(reshape2)
library(scoringRules)
source("Code/helper_plots.R")
source("Code/helper_calculators.R")
source("Code/RE_dose_response_MCMC.R")
source("Code/dirichlet_MCMC.R")
# read in data
source("Code/tox21_prep_data.R")
```
The last line prepared a series of objects representing the data, including the
observed dose response $y_i$, the concentration for the observed dose response
$Cx$, and a list to track which entries are replicates.  Null values are present
in the data to indicate an anomalous response, such as cell death.  Some
chemicals have 3 replicates and some have 6.
```{r prepped input}
message("Observed response")
head(y_i)
message("Concentration")
head(Cx)
message("Replicates by index")
head(replicate_sets)
```
### Fitting Dose Responses A random effect model is fit to each chemical indexed
by $i$, $$  R_i(c) =f_i(c|a_i,\theta_i,\beta_i, u_{ij}, v_{ij}) =
\frac{a_i+u_{ij}}{1+\left(\frac{\theta_i}{c}\right)^{\beta_i}} + v_{ij} +
\epsilon_{ijc} $$ The MCMC script uses standard methods and an alternative
script is available for Nimble.
```{r RE-MCMC}
re_iter = 2.5e4 #25,000 iterations
set.seed(102)
# fit random effects model
re_chains = RE_MCMC_fit(y_i, Cx, replicate_sets, n_iter = re_iter)
```
The beginning 5000 iterations of the chains are removed and the remaining
iterations thinned.
```{r}
re_par_list = pull_parameters(re_chains)
```
The slope parameters are of particular interest, since they are used next for
clustering.  It is hypothesized that the slope parameter is related to the mode
of action for a chemical, and that chemicals with a similar mode of action
follow concentration addition.  A Dirichlet Process model is used for
clustering.
### Clustering Slopes
```{r}
clust_iter =3.5e4  #3.5e4
set.seed(131)
cluster_chain = DP_MCMC_fit((re_par_list$slope_params[1:18]), n_iter = clust_iter)
beepr::beep()
```
The MCMC procedure produces a chain of possible clusterings, with more likely
clusterings occuring more often.  We extract the top 20 most frequent (likely)
clusters.
```{r}
# check what top assignments are
n_top = 20
clust_centers_w_prob = cluster_centers(cluster_chain, n_top = n_top)
```
A helper function plots what the top clusters look like.
```{r}
# visualize clusters
visualize_clusters_blocks(re_par_list$slope_params, clust_centers_w_prob)
```
Next we create the lists that contain all the necessary parameters for our model
with Reflected GCA and Dirichlet Process clusters (RGD).  Note that one of the
parameters is the probability of each cluster, which is computed from the counts
shown in the visualization on the right (using $p_i = c_i / \sum_{j=1}^{20}
c_j$)
```{r}
cluster_prob = clust_centers_w_prob$probs
centers = clust_centers_w_prob$centers
cent_sd = clust_centers_w_prob$center_sd
cluster_assign = clust_centers_w_prob$assign
DP_par_list = list("centers" = centers,
                   "cluster_assign" = cluster_assign,
                   "cent_sd" = cent_sd,
                   "cluster_prob" = cluster_prob)
tot_par_list = c(re_par_list, DP_par_list)
```
We make the parameter lists for the simpler Independent Action (IA) and regular
GCA models.
```{r}
# fixed assignments for GCA and IA
GCA_assign = 1
names(GCA_assign) = do.call(paste, as.list(rep(1, n_chems)))
DP_2par_list = list("centers" = matrix(rep(1, n_chems),nrow=1),
                    "cluster_assign" = GCA_assign,
                    "cent_sd" = matrix(rep(0, n_chems), nrow=1),
                    "cluster_prob" = 1)
GCA_par_list = c(re_2par_list, DP_2par_list)
# for IA, allow different slopes
IA_assign =1
names(IA_assign) = do.call(paste, as.list( 1:n_chems))
DP_2parIA_list = list("centers" = matrix(re_par_summary$slope_params, nrow = 1),#matrix(rep(1, n_chems),nrow=1),
                      "cluster_assign" = IA_assign,
                      "cent_sd" = matrix(rep(0, n_chems), nrow=1),
                      "cluster_prob" = 1)
IA_par_list = c(re_par_list, DP_2parIA_list)
beepr::beep(2)
```
One of the special cases we test is to use the Reflected GCA but without the
clustering.  Hence we create one large cluster (GCA) and keep the fitted slopes.
Regular GCA must set the slope values to 1.
```{r}
# For RGCA no DP, allow different slopes
GCA_assign = 1
names(GCA_assign) = do.call(paste, as.list(rep(1, n_chems)))
RDP_2par_list = list("centers" = matrix(re_par_summary$slope_params, nrow = 1),
                     "cluster_assign" = GCA_assign,
                     "cent_sd" = matrix(rep(0, n_chems), nrow=1),
                     "cluster_prob" = 1)
RGCA_par_list = c(re_2par_list, RDP_2par_list)
```
Given the parameter lists, we can instantiate the calculators that compute the
mixture response given the individual concentrations.  For the methods that use
clustering, we sample with replacement from the set of possible clusters with
weighted probability.  The chemical parameters are sampled as well, conditional
on the clustering (mainly the slope parameters depend on the clustering)
```{r}
set.seed(1026)
n_bootstraps = 100
samp_idx = sample(1:n_top, size = n_bootstraps, prob =cluster_prob, replace = T)
sampled_mix_funs = sapply(samp_idx, FUN = function(x) create_mix_calc(x, tot_par_list, add_RE = T))
sampled_mix_funs_GCA = sapply(rep(1, n_bootstraps), FUN = function(x) create_mix_calc(x, GCA_par_list, add_RE = F))
sampled_mix_funs_IA = sapply(rep(1, n_bootstraps), FUN = function(x) create_mix_calc(x, IA_par_list, add_RE = F))
# additional comparisons:
# use unit slopes, so no RGCA
sampled_mix_funs_GCA_noR = sapply(samp_idx, FUN = function(x) create_mix_calc(x, tot_par_list, add_RE = F, unit_slopes=T))
# only use the top clustering, no DP
sampled_mix_funs_noDP = sapply(rep(1, n_bootstraps), FUN = function(x) create_mix_calc(x, RGCA_par_list, add_RE = F))
bootstrap_calc_list = list("RGD" = sampled_mix_funs,
                           "G" = sampled_mix_funs_GCA,
                           "I" = sampled_mix_funs_IA,
                           "DG" = sampled_mix_funs_GCA_noR,
                           "RG" = sampled_mix_funs_noDP)
```
###Predicting Mixture responses
Before proceeding with actual mixtures, we can check that sham mixtures with
only one chemical are correctly modeled.
```{r}
plot_dummy_mixture(Cx, y_i, tot_par_list, replicate_sets,
                   bootstrap_calc_list, test_idx =5)
```
Now we can proceed to predicting with the actual data.  The sets of indices are
chosen based on the descriptions of the mixtures. This part is solving the
two-step model with the bootstrapped calculators built on our fitted parameters
and clustering, and can take a few minutes per mixture
```{r}
# all mixtures where the Estrogen Receptor agonists are at 4x their fitted EC50
set_4x = c( 1, 12, 20, 29, 43, 52, 55, 57)
# mixtures from set_4x plus all other mixtures with all chemicals or all
# Androgen receptor agonists present
set_1 = c(1, 5, 10, 12, 20, 25, 29, 30, 31, 32, 43, 45, 50, 52, 55, 57, 62)
# a single index for testing
mix_idx = 1#16, 8, 27
# a sampling of mixtures with only two components
binary_mixes = c(8, 34, 37, 61)
# a sampling of chemicals with 3-5 components
small_mixes = c(6, 26, 47 )
# just plot one index for demonstration
score_matrix = plot_mixture_response(mix_idx, mix_df, mix_conc_df, mix_guide,
                                     bootstrap_calc_list)
beepr::beep(2)
```


Finally, we format the score matrix and plot the results of the score comparisons.

```{r}
score_df = data.frame(score_matrix)
# drop rows of 0, in case some mixtures were skipped
score_df = score_df[apply(score_df, MARGIN=1, FUN = function(rx) any(rx>0)),]
names(score_df) = c("Mix id", paste(names(bootstrap_calc_list), c("LLH")),
                    paste(names(bootstrap_calc_list), c("MSE")),
                    paste(names(bootstrap_calc_list), c("CRPS")))

## Violin Plot for Scores ####
plot_scores(score_df[subindx,], bootstrap_calc_list)

```

