---
title: "Simple Application of RGCA"
author: "Daniel Zilber"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RGCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RGCA)
```

## Simple Usage of RGCA

While the method could be generalize to a variety of smooth, monotone dose response functions, our package is designed around the Hill function,
$$f(x |a,b,c) =
\frac{a}{1+\left(\frac{b}{x}\right)^{c}}$$
The parameters are the sill (a), the EC50 (b), and the slope (c). Give these parameters and a cluster assignment vector, RGCA can create a calculator that predicts the mixture response given an input dose vector $(x_1,...,x_n)$. In the example below, there are three chemicals with known Hill parameters.

```{r}
n_chems <- 3
sills <- c(3, 5, 4)
ec50_vec <- c(1, 0.75, 2.4)
slopes <- c(0.5, 1.1, 2.0)
# Rmax is used to scale IA across clusters, can copy sills
param_matrix <- as.matrix(cbind("a" = sills,
                                "b" = ec50_vec,
                                "c" = slopes,
                                "max_R" = sills))
# specify both chems in cluster 1 of 1
cluster_assign_vec <- c(1, 2, 1)
# create a calculator to predict response given concentration
mix_pred <- mix_function_generator(param_matrix, cluster_assign_vec)
```

To create a mixture, we need to specify the concentration of each chemical.

```{r pressure}
# generate mix concentrations:  each row of the matrix is one dose of the mix
n_samps <- 30
# equipotent mixture: concentrations scaled by EC50
equipot_conc_matrix <- matrix(0, nrow = n_samps, ncol = n_chems)
# equimolar mixture: equal concentration of all chemicals
equimol_conc_matrix <- matrix(0, nrow = n_samps, ncol = n_chems)
# generate concentrations on the log scale
for (chem_idx in 1:n_chems) {
  equipot_conc_matrix[, chem_idx] <-
    ec50_vec[chem_idx] / (10^seq(2, -1, length.out = n_samps))
  equimol_conc_matrix[, chem_idx] <-
    1 / (10^seq(3, -1, length.out = n_samps))
}

#create the mixture concentration vector for plotting
equipot_conc <- rowSums(equipot_conc_matrix)
equimol_conc <- rowSums(equimol_conc_matrix)

# Apply the pediction function to the concentrations of interest
pred_equipot <- apply(equipot_conc_matrix,
                      MARGIN = 1,
                      FUN = function(x) mix_pred(x))
pred_equimol <- apply(equimol_conc_matrix,
                      MARGIN = 1,
                      FUN = function(x) mix_pred(x))

```

Now we can plot the two predicted mixture responses.


```{r}
plot(equimol_conc, pred_equimol, log = "x", ylim = c(0, 5))
points(equipot_conc, pred_equimol, col = 2)
```


## Fitting Dose Response Curves with Nimble

In the example above, we assume the parameters are known.  If you have raw data for the individual chemical dose responses and want to fit the random effect model that we describe in our manuscript, the RGCA package has a built-in fitting routing based on Nimble.  Three pieces of data are needed:  the responses, the doses, and a list of the replicates.  
